---
title: "Sample size for a trial analysed with non-responder imputation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample size for a trial analysed with non-responder imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4, fig.align = "center")
```

```{r setup}
library(ssNRI)
```

## Two analyses, two estimands

A trial with a binary endpoint and dropout can be analysed in two ways. A
complete case analysis compares the response probabilities among those who
complete. Non-responder imputation counts a dropout as a non-responder and
compares

$$p_{j,\mathrm{NRI}} = \Pr(R_{j} = 1, D_{j} = 0),$$

where $R_{j}$ is the latent response indicator and $D_{j}$ the dropout
indicator in group $j$. These are different quantities, and the difference is
not a nuisance: the effect that an NRI analysis is powered to detect,
$\delta_{\mathrm{NRI}} = p_{1,\mathrm{NRI}} - p_{2,\mathrm{NRI}}$, is not the
effect a complete case analysis is powered to detect.

The usual practice is to compute a sample size for a trial without dropout and
inflate it by $1 / (1 - \omega)$. That adjustment restores the number of
completers, which is what a complete case analysis needs. It does nothing about
the change of estimand that NRI brings, so it is not the right adjustment when
NRI is planned.

`ssNRI` models $R_{j}$ and $D_{j}$ jointly as a bivariate Bernoulli distribution,
which makes both estimands explicit and gives a closed-form sample size for
each.

## Two ways to state the same assumption

The joint distribution needs one more parameter than the two marginal
probabilities $p_{j}$ and $\omega_{j}$. It can be given as a correlation
$\rho_{j}$, or as

$$\gamma_{j} = \Pr(R_{j} = 1 \mid D_{j} = 1),$$

the probability that a dropout would have responded. The two are in one to one
correspondence,

$$\gamma_{j} = p_{j} + \rho_{j} \sqrt{\frac{p_{j} (1 - p_{j}) (1 - \omega_{j})}{\omega_{j}}},
\qquad
\rho_{j} = (\gamma_{j} - p_{j}) \sqrt{\frac{\omega_{j}}{p_{j} (1 - p_{j}) (1 - \omega_{j})}}.$$

```{r gamma-conversion}
rho_to_gamma(p = 0.45, omega = 0.10, rho = 0)
rho_to_gamma(p = 0.45, omega = 0.10, rho = c(-0.2, 0.2))
gamma_to_rho(p = 0.45, omega = 0.10, gamma = 0.60)
```

Independence gives $\gamma_{j} = p_{j}$: dropouts respond at the same rate as
the group as a whole. The NRI response probability is
$p_{j,\mathrm{NRI}} = p_{j} - \gamma_{j} \omega_{j}$, so $\gamma_{j}$ is exactly
the fraction of dropouts that non-responder imputation misclassifies, and

$$\delta_{\mathrm{NRI}} = (p_{1} - \gamma_{1} \omega_{1}) - (p_{2} - \gamma_{2} \omega_{2}).$$

Because the four joint probabilities must be non-negative, neither parameter is
free over its whole nominal range. On the $\gamma$ scale the restriction is
plain arithmetic,

$$\max \left\{ 0, \frac{p_{j} + \omega_{j} - 1}{\omega_{j}} \right\}
\leq \gamma_{j} \leq
\min \left\{ 1, \frac{p_{j}}{\omega_{j}} \right\},$$

which is the requirement that the four joint probabilities be non-negative, and
it maps onto the Frechet-Prentice bounds on the correlation.

```{r gamma-bounds}
gamma_bounds(p = 0.45, omega = 0.10)
rho_bounds(p = 0.45, omega = 0.10)
```

The bounds are 0 and 1 here, and they usually are: $\gamma_{j}$ is an ordinary
conditional probability and the restriction binds only once the dropout
probability exceeds $p_{j}$ or $1 - p_{j}$. What is worth looking at is the
exchange rate between the two scales.

```{r gamma-bounds-plot}
plot(gamma_bounds(p = 0.45, omega = 0.10))
```

Each segment spans the attainable correlations at one dropout probability, and
all four span the whole unit interval in $\gamma_{j}$. The slope is what
differs. At $\omega_{j} = 0.05$ a correlation of $0.25$ already means that every
dropout would have responded, while at $\omega_{j} = 0.30$ the same statement
needs a correlation of about $0.72$. An assumption that reads as mild on the
correlation scale can be a strong claim about what the dropouts would have done,
which is the reason to elicit it on the $\gamma$ scale.

## A worked design

The phase 3 cytisine trial of Dogar et al. (2020) planned for abstinence
probabilities of 0.47 and 0.41, a dropout probability of 0.10, equal
allocation, a one-sided 0.025 level and 80 percent power, and analysed the
primary endpoint under NRI. The simple inflation method gives the sample size
the trial enrolled.

```{r design-cc}
p1 <- 0.47
p2 <- 0.41
omega <- 0.10

cc <- sample_size_cc(p1 = p1, p2 = p2, omega1 = omega, omega2 = omega,
                     r = 1, alpha = 0.025, target_power = 0.8)
cc$n_total_adjust
```

The NRI formula needs the association parameter as well. The correlation is not
free, so a natural sensitivity range runs from the lowest to the highest value
both groups admit.

```{r design-nri}
b1 <- rho_bounds(p = p1, omega = omega)
b2 <- rho_bounds(p = p2, omega = omega)
rho_range <- c(L = max(b1$rho_lower, b2$rho_lower),
               "0" = 0,
               U = min(b1$rho_upper, b2$rho_upper))

design <- do.call(rbind, lapply(names(rho_range), function(nm) {
  rho <- rho_range[[nm]]
  ss <- sample_size_nri(p1 = p1, p2 = p2, omega1 = omega, omega2 = omega,
                        rho1 = rho, rho2 = rho,
                        r = 1, alpha = 0.025, target_power = 0.8)
  data.frame(
    rho_label = nm,
    rho = rho,
    gamma1 = rho_to_gamma(p = p1, omega = omega, rho = rho),
    gamma2 = rho_to_gamma(p = p2, omega = omega, rho = rho),
    delta_NRI = ss$effect_NRI,
    N_CC = cc$n_total_adjust,
    N_NRI = ss$n_total,
    RE = cc$n_total_adjust / ss$n_total
  )
}))
design
```

The $\gamma$ columns say what the two extremes of the correlation mean. At the
lower bound no dropout in the control group would have responded; at the upper
bound every dropout in the treatment group would have. The full-data effect is
0.06 throughout, while the effect an NRI analysis is powered against is smaller,
and the sample size the simple inflation method gives falls short of the sample
size NRI needs at every value of the correlation considered here.

The `plot()` method sweeps the same comparison across the feasible range rather
than at three points.

```{r design-plot}
ss <- sample_size_nri(p1 = p1, p2 = p2, omega1 = omega, omega2 = omega,
                      rho1 = 0, rho2 = 0, r = 1, alpha = 0.025,
                      target_power = 0.8)
plot(ss, vary = "rho")
```

## Exact operating characteristics

Nothing above is checked by simulation. `power_nri()` and `power_cc()`
enumerate the joint distribution of the observed outcomes, so the power and the
type I error rate they return are exact up to floating point.

```{r exact-power}
planned <- power_nri(n1 = 1194, n2 = 1194, p1 = p1, p2 = p2,
                     omega1 = omega, omega2 = omega, alpha = 0.025)
required <- power_nri(n1 = 1287, n2 = 1287, p1 = p1, p2 = p2,
                      omega1 = omega, omega2 = omega, alpha = 0.025)
c(planned = planned$power, required = required$power)
```

A design built by inflating a dropout-free sample size therefore attains about
77 percent power against the effect it was designed to detect, not the 80
percent intended.

`power_cc()` does the same for a complete case analysis. A direct enumeration
over completer counts and responder counts is $O(N^{4})$; `power_cc()` uses the
fact that the pooled-variance $Z$ statistic increases in the group 1 responder
count wherever it is positive, so the rejection region is an upper set and the
inner double sum collapses to a binomial tail probability at one threshold per
control-group count. That is a regrouping of the same finite sum, so the
returned values remain exact.

```{r exact-power-cc}
power_cc(n1 = 60, n2 = 60, p1 = 0.6, p2 = 0.4,
         omega1 = 0.15, omega2 = 0.15, alpha = 0.025)
```

## What a historical rate does not tell you

Design parameters usually come from a published trial, and a published rate is
the output of an analysis rule rather than a statement about what the dropouts'
outcomes were. The 0.41 above came from an earlier trial that counted every
patient without verified abstinence as a non-abstainer. That rate is
$\Pr(R = 1, D = 0)$, not $\Pr(R = 1)$, so the latent probability is identified
only within an interval.

```{r latent-bounds}
lb <- latent_bounds(p_obs = 0.41, N_randomized = 620, N_dropout = 62)
lb
```

The interval always has width $\omega$; only its position depends on the rule
the historical trial applied. Counting dropouts as responders, as a best case
analysis does in the treatment group, shifts it down by $\omega$; a complete
case rate is first placed back onto the same scale.

```{r latent-bounds-rules}
rules <- expand.grid(method = c("nri", "cc", "bc", "wc"),
                     group = c("treatment", "control"),
                     stringsAsFactors = FALSE)
intervals <- do.call(rbind, Map(function(m, g) {
  b <- latent_bounds(p_obs = 0.41, N_randomized = 620, N_dropout = 62,
                     method = m, group = g)
  data.frame(method = m, group = g, p_lower = b$p_lower, p_upper = b$p_upper)
}, rules$method, rules$group))
rownames(intervals) <- NULL
intervals
```

Each candidate value in the interval implies a different correlation, which
`infer_rho()` returns. Setting a joint cell to zero, as is sometimes done to
read a latent probability off a best or worst case rate, selects one endpoint of
the interval rather than identifying a point.

```{r infer-rho}
for (p in c(lb$p_lower, lb$p_midpoint, lb$p_upper)) {
  cat(sprintf("p = %.3f  ->  rho = %+.4f\n", p,
              infer_rho(p_obs = 0.41, omega = lb$omega, p = p)$rho))
}
```

Planning should carry the interval rather than a single recovered value.

## Reproducing the article

`inst/scripts/` regenerates the figures and tables of the main text and
`inst/sup_info/` those of the supplementary material. Each is split into a
computation stage that writes intermediate results to `data/*.rds` and a
rendering stage that reads those files and writes `results/*.eps`,
`results/*.pdf` and `results/*.tex`. `inst/scripts/number_check.R` re-extracts
every numerical value quoted in the article from the regenerated `.rds` files,
so that a number transcribed into the manuscript can be checked against its
source.

## Reference

Dogar O, Keding A, Gabe R, et al. Cytisine for smoking cessation in patients
with tuberculosis: a multicentre, randomised, double-blind, placebo-controlled
phase 3 trial. *Lancet Global Health* 2020; 8: e1408-e1417.
